Visual Diagnostics of an Explainer Model -- Tools for the Assessment of LIME Explanations

### 2021-01-04 paper.Rnw
finished creating supplemental material document
updated text to account for supporting info document

% Citation information
\jnlcitation{
\cname{\author{Goode K.}, \author{H. Hofmann}, }
\cyear{2020},
\ctitle{Visual Diagnostics of an  Explainer Model -- Tools for the Assessment of LIME Explanations},
\cjournal{Stat Anal Data Min: The ASA Data Sci Journal},
\cvol{volume, number and page}.}

@@

@
\caption{Hypothetical scenarios depicting the effect of different weights on LIME. (Left) An example of an explainer model with appropriate weights to provide a faithful approximation of the complex model. (Right) An example of an explainer model for the same prediction but with different weights that results in an explanation unfaithful to the complex model.  }
\label{fig:figure-01}
\end{figure*}

In the field of statistics, there are two main uses for models: inference and prediction. Machine learning models  have proven to perform well in problems with the latter objective , but the accuracy of many machine learning models comes at the cost of interpretability due to their algorithmic complexity (hence the phrase "black-box models"). Model interpretability allows for the understanding and assessment of how a model produces predictions. The lack of the ability to understand and assess a model makes it difficult to trust the model, especially in areas with high stakes decisions such as the medical and forensics sciences. The increased use of machine learning models in applications and the introduction of the General Data Protection Regulation (GDPR) in 2018 \citep{goodman:2016} has resulted in a dramatic increase in explainable machine learning research, which focuses on developing ways to explain output from machine learning algorithms.

@@

\autoref{fig:figure-01} demonstrates a simple example in which the explanation quality varies drastically when different weights are used. Other concerns with LIME have been raised in the literature: \citet{laugel:2018} and \citet{molnar:2019} discuss the difficulty specifying a local region with LIME due to both  an unclear definition of what is a "local region" and how to apply LIME to achieve an appropriate local region. \citet{laugel:2018} also present examples of LIME explanations that are clearly global and not local. A different concern about LIME is raised by \citet{alvarezmelis:2018}  pertaining to the robustness of explanations from LIME and other explainer models : they find that even small changes in predictor variables can lead to very different LIME explanations. Even the original authors of LIME, \citet{ribeiro:2016}, acknowledge that if a linear model is used as the explainer, LIME relies on a linear approximation of the explainer model to the complex model and state "if the underlying model is highly non-linear even in the locality of the prediction, there may not be a faithful explanation".

@@

@
\caption{Plots of $x_2$ versus $x_1$ from the \data \  training (left) and testing (right) sets introduced in Section \ref{diagnostics}  with the true classification boundary  shown as a solid black line.  In the testing set, the open black circles identify cases misclassified by the random forest, and the diamond indicates a prediction of interest. }
\label{fig:figure-02}
\end{figure*}

@@

To demonstrate the visual diagnostics, we generate an example dataset that will be referred to as the \data. The \data \ contains \Sexpr{dim(sine_data)[1]} observations with  features of $x_1$ and $x_2$ independently sampled from Unif(\Sexpr{min}, \Sexpr{max}) distributions and $x_3$  sampled from a $\mbox{N}(0,1)$ distribution. A binary response variable $y$ is created using a  sine curve  such that
\begin{eqnarray}\label{eq:data}
  y=\begin{cases}
  \mbox{blue} & \mbox{ if } x'_2 > \sin\left(x'_1\right) \\
  \mbox{red} & \mbox{ if } x'_2 \le \sin\left(x'_1\right) \ .
  \end{cases}
\end{eqnarray}
where $x'_1=x_1\cos(\theta)-x_2\sin(\theta)$,  $x'_2=x_1\sin(\theta)+x_2\cos(\theta)$, and  $\theta=-0.9$. \sout{Then $y$ is defined as
\begin{eqnarray}\label{eq:data}
  y=\begin{cases}
  \mbox{blue} & \mbox{ if } x'_2 > \sin\left(x'_1\right) \\
  \mbox{red} & \mbox{ if } x'_2 \le \sin\left(x'_1\right) \ .
  \end{cases}
\end{eqnarray}}
Note that due to the creation of $y$ in this manner, $y$ is dependent on $x_1$ and $x_2$ and independent of $x_3$. The dataset is randomly divided into training and testing sets of \Sexpr{dim(sine_data_train)[1]} and \Sexpr{dim(sine_data_test)[1]} observations, respectively. A random forest is fit to the training set using the R package \emph{randomForest} (version \Sexpr{packageVersion("randomForest")}) \citep{liaw:2002} with the default settings and is applied to the test set to obtain predictions.

@@

When interpreting feature heatmaps, horizontal lines of the same color represent the ideal situation where the explanations are consistent and do not depend on the tuning parameter values but do depend on the location of the observation in the feature space. Vertical lines of color represent explanations that are dependent on the tuning parameter values and may not be local since the same feature is chosen regardless of the location of the observation in the feature space. See Appendix \ref{feat-heat-ex} for hypothetical examples depicting feature heatmaps with the extreme cases of horizontal and vertical lines.

@@

@
\caption{Example bullet signatures. The  signatures are from  the same land and therefore have very similar patterns. }
\label{fig:figure-07}
\end{figure}

@@

To provide a more detailed view of explanations obtained using different tuning parameter values, we consider explanation scatterplots for one prediction of interest  referred to as case NM (a known non-match) shown in Figure~\ref{fig:figure-11}  along with the visual  explanation plots from the \emph{lime} R package. Plots for 4-equal-bins and 4-quantile-bins are included to depict tuning parameter values with good (4-equal-bins) and mediocre (4-quantile-bins) performance based on the assessment metric plot. %\kge{Also, recall that 4-quantile-bins are the default option in the \emph{lime} R package.} 
Explanation scatterplots for a known match, referred to as case M, and for the kernel density simulation method applied to both M and NM are included in Appendix~\ref{bullets-plus}.

@@

\begin{figure*}[!thp]
<<figure-11, out.width = '6.25in', fig.width = 13, fig.height = 8.5, message = FALSE>>=

@@

@@

@
\caption{Plots of LIME explanations (first row) and explanation scatterplots (second row) for case NM in the bullet test data for two tuning parameter values: 4-quantile-bins (first column) and 4-equal-bins (second column). }
\label{fig:figure-11}
\end{figure*}

@@

The code used to produce the manuscript and the data from examples in Section \ref{application} are available in the following GitHub Repository: \href{https://github.com/goodekat/LIME-diagnostics-paper}{https://github.com/goodekat/LIME-diagnostics-paper}.

>>>>

% Citation information
\jnlcitation{
\cname{\author{Goode K.}, \author{H. Hofmann}},
\cyear{2020},
\ctitle{Visual Diagnostics of an Explainer Model -- Tools for the Assessment of LIME Explanations},
\cjournal{Stat Anal Data Min: The ASA Data Sci Journal},
\cvol{volume, number and page}.}

@@

@
\caption{Hypothetical scenarios depicting the effect of different weights on LIME. (Left) An example of an explainer model with appropriate weights to provide a faithful approximation of the complex model. (Right) An example of an explainer model for the same prediction but with different weights that results in an explanation unfaithful to the complex model.}
\label{fig:figure-01}
\end{figure*}

In the field of statistics, there are two main uses for models: inference and prediction. Machine learning models have proven to perform well in problems with the latter objective, but the accuracy of many machine learning models comes at the cost of interpretability due to their algorithmic complexity (hence the phrase "black-box models"). Model interpretability allows for the understanding and assessment of how a model produces predictions. The lack of the ability to understand and assess a model makes it difficult to trust the model, especially in areas with high stakes decisions such as the medical and forensics sciences. The increased use of machine learning models in applications and the introduction of the General Data Protection Regulation (GDPR) in 2018 \citep{goodman:2016} has resulted in a dramatic increase in explainable machine learning research, which focuses on developing ways to explain output from machine learning algorithms.

@@

\autoref{fig:figure-01} demonstrates a simple example in which the explanation quality varies drastically when different weights are used. Other concerns with LIME have been raised in the literature: \citet{laugel:2018} and \citet{molnar:2019} discuss the difficulty specifying a local region with LIME due to both an unclear definition of what is a "local region" and how to apply LIME to achieve an appropriate local region. \citet{laugel:2018} also present examples of LIME explanations that are clearly global and not local. A different concern about LIME is raised by \citet{alvarezmelis:2018} pertaining to the robustness of explanations from LIME and other explainer models: they find that even small changes in predictor variables can lead to very different LIME explanations. Even the original authors of LIME, \citet{ribeiro:2016}, acknowledge that if a linear model is used as the explainer, LIME relies on a linear approximation of the explainer model to the complex model and state "if the underlying model is highly non-linear even in the locality of the prediction, there may not be a faithful explanation".

@@

@
\caption{Plots of $x_2$ versus $x_1$ from the \data \ training (left) and testing (right) sets introduced in Section \ref{diagnostics} with the true classification boundary shown as a solid black line. In the testing set, the open black circles identify cases misclassified by the random forest, and the diamond indicates a prediction of interest.}
\label{fig:figure-02}
\end{figure*}

@@

To demonstrate the visual diagnostics, we generate an example dataset that will be referred to as the \data. The \data \ contains \Sexpr{dim(sine_data)[1]} observations with  features of $x_1$ and $x_2$ independently sampled from Unif(\Sexpr{min}, \Sexpr{max}) distributions and $x_3$ sampled from a $\mbox{N}(0,1)$ distribution. A binary response variable $y$ is created using a sine curve such that
\begin{eqnarray}\label{eq:data}
  y=\begin{cases}
  \mbox{blue} & \mbox{ if } x'_2 > \sin\left(x'_1\right) \\
  \mbox{red} & \mbox{ if } x'_2 \le \sin\left(x'_1\right) \ .
  \end{cases}
\end{eqnarray}
where $x'_1=x_1\cos(\theta)-x_2\sin(\theta)$,  $x'_2=x_1\sin(\theta)+x_2\cos(\theta)$, and  $\theta=-0.9$. Note that due to the creation of $y$ in this manner, $y$ is dependent on $x_1$ and $x_2$ and independent of $x_3$. The dataset is randomly divided into training and testing sets of \Sexpr{dim(sine_data_train)[1]} and \Sexpr{dim(sine_data_test)[1]} observations, respectively. A random forest is fit to the training set using the R package \emph{randomForest} (version \Sexpr{packageVersion("randomForest")}) \citep{liaw:2002} with the default settings and is applied to the test set to obtain predictions.

@@

When interpreting feature heatmaps, horizontal lines of the same color represent the ideal situation where the explanations are consistent and do not depend on the tuning parameter values but do depend on the location of the observation in the feature space. Vertical lines of color represent explanations that are dependent on the tuning parameter values and may not be local since the same feature is chosen regardless of the location of the observation in the feature space. See \kge{the supporting information (Figure S1)} \sout{Appendix \ref{feat-heat-ex}} for hypothetical examples depicting \kge{the extreme cases of} feature heatmaps\kge{.} \sout{with the extreme cases of horizontal and vertical lines.}

@@

@
\caption{Example bullet signatures. The signatures are from the same land and therefore have very similar patterns.}
\label{fig:figure-07}
\end{figure}

@@

To provide a more detailed view of explanations obtained using different tuning parameter values, we consider explanation scatterplots for one prediction of interest referred to as case NM (a known non-match) shown in Figure~\ref{fig:figure-11} along with the visual explanation plots from the \emph{lime} R package. Plots for 4-equal-bins and 4-quantile-bins are included to depict tuning parameter values with good (4-equal-bins) and mediocre (4-quantile-bins) performance based on the assessment metric plot. \kge{Additional examples of explanation scatterplots including explanations for a known match and explanation produced using the kernel density simulation method are included in the  the supporting information (Figures S2 to S4).} \sout{Explanation scatterplots for a known match, referred to as case M, and for the kernel density simulation method applied to both M and NM are included in Appendix~\ref{bullets-plus}.}

@@

\begin{figure*}[!thp]
<<figure-11, out.width = '6.25in', fig.width = 13, fig.height = 9, message = FALSE>>=

@@

# Save the cleaned explanations to use with the supporting information
if(!file.exists("data/bullet-explain-perms-clean.rds")) {
 saveRDS(bullet_explain_perms_clean, "data/bullet-explain-perms-clean.rds") 
}

@@

@
\caption{Plots of LIME explanations (first row) and explanation scatterplots (second row) for case NM in the bullet test data for two tuning parameter values: 4-quantile-bins (first column) and 4-equal-bins (second column).}
\label{fig:figure-11}
\end{figure*}

@@

\kge{Additional example visual diagnostics may be found online in the Supporting Information section at the end of this manuscript.} The code used to produce the manuscript and the data from examples in Section \ref{application} are available in the following GitHub Repository: \href{https://github.com/goodekat/LIME-diagnostics-paper}{https://github.com/goodekat/LIME-diagnostics-paper}.

### 2021-01-04 paper.tex
updated text to account for supporting info document

\end{knitrout}
\caption{Hypothetical scenarios depicting the effect of different weights on LIME. (Left) An example of an explainer model with appropriate weights to provide a faithful approximation of the complex model. (Right) An example of an explainer model for the same prediction but with different weights that results in an explanation unfaithful to the complex model.  }
\label{fig:figure-01}
\end{figure*}

In the field of statistics, there are two main uses for models: inference and prediction. Machine learning models  have proven to perform well in problems with the latter objective , but the accuracy of many machine learning models comes at the cost of interpretability due to their algorithmic complexity (hence the phrase "black-box models"). Model interpretability allows for the understanding and assessment of how a model produces predictions. The lack of the ability to understand and assess a model makes it difficult to trust the model, especially in areas with high stakes decisions such as the medical and forensics sciences. The increased use of machine learning models in applications and the introduction of the General Data Protection Regulation (GDPR) in 2018 \citep{goodman:2016} has resulted in a dramatic increase in explainable machine learning research, which focuses on developing ways to explain output from machine learning algorithms.

@@

\end{knitrout}
\caption{Plots of $x_2$ versus $x_1$ from the \data \  training (left) and testing (right) sets introduced in Section \ref{diagnostics}  with the true classification boundary  shown as a solid black line.  In the testing set, the open black circles identify cases misclassified by the random forest, and the diamond indicates a prediction of interest. }
\label{fig:figure-02}
\end{figure*}

@@

\end{knitrout}
\caption{Example bullet signatures. The  signatures are from  the same land and therefore have very similar patterns. }
\label{fig:figure-07}
\end{figure}

@@

\end{knitrout}
\caption{Plots of LIME explanations (first row) and explanation scatterplots (second row) for case NM in the bullet test data for two tuning parameter values: 4-quantile-bins (first column) and 4-equal-bins (second column). }
\label{fig:figure-11}
\end{figure*}

>>>>

\end{knitrout}
\caption{Hypothetical scenarios depicting the effect of different weights on LIME. (Left) An example of an explainer model with appropriate weights to provide a faithful approximation of the complex model. (Right) An example of an explainer model for the same prediction but with different weights that results in an explanation unfaithful to the complex model.}
\label{fig:figure-01}
\end{figure*}

In the field of statistics, there are two main uses for models: inference and prediction. Machine learning models have proven to perform well in problems with the latter objective, but the accuracy of many machine learning models comes at the cost of interpretability due to their algorithmic complexity (hence the phrase "black-box models"). Model interpretability allows for the understanding and assessment of how a model produces predictions. The lack of the ability to understand and assess a model makes it difficult to trust the model, especially in areas with high stakes decisions such as the medical and forensics sciences. The increased use of machine learning models in applications and the introduction of the General Data Protection Regulation (GDPR) in 2018 \citep{goodman:2016} has resulted in a dramatic increase in explainable machine learning research, which focuses on developing ways to explain output from machine learning algorithms.

@@

\end{knitrout}
\caption{Plots of $x_2$ versus $x_1$ from the \data \ training (left) and testing (right) sets introduced in Section \ref{diagnostics} with the true classification boundary shown as a solid black line. In the testing set, the open black circles identify cases misclassified by the random forest, and the diamond indicates a prediction of interest.}
\label{fig:figure-02}
\end{figure*}

@@

\end{knitrout}
\caption{Example bullet signatures. The signatures are from the same land and therefore have very similar patterns.}
\label{fig:figure-07}
\end{figure}

@@

\end{knitrout}
\caption{Plots of LIME explanations (first row) and explanation scatterplots (second row) for case NM in the bullet test data for two tuning parameter values: 4-quantile-bins (first column) and 4-equal-bins (second column).}
\label{fig:figure-11}
\end{figure*}
